Transit Delay Patterns in 2024 Toronto: A Comparative Analysis of Geographic, Modal, and Temporal Factors

### 2024-12-05 paper/paper.qmd
final paper finished

---
title: ""
author: 
  - Jerry Xia
thanks: "Code and data are available at: <https://github.com/Jerryx2020/TTC_Delay_Analysis>"
date: today
date-format: long
abstract: ""
format: pdf
number-sections: true
toc: true 
bibliography: references.bib
---

@@

library(tidyverse)
library(ggplot2)
```

>>>>

---
title: "Transit Delay Patterns in 2024 Toronto: A Comparative Analysis of Geographic, Modal, and Temporal Factors"
subtitle: "Uncovering Operational Challenges and Opportunities for Improving Toronto’s Transit System"
author: 
  - Jerry Xia
thanks: "Code and data are available at: <https://github.com/Jerryx2020/TTC_Delay_Analysis>"
date: "December 3, 2024"
date-format: long
abstract: "This study investigates delays in the Toronto Transit Commission (TTC), focusing on buses, subways, and streetcars, to understand factors affecting transit reliability. Using real-world data from Open Data Toronto and a random forest model, it identifies key predictors of delays, including transit mode, geographic location, and temporal patterns. The analysis reveals that buses experience the most frequent and longest delays, particularly in urban areas like Dundas, while suburban locations such as Yorkdale face fewer disruptions. These findings highlight the operational challenges unique to transit modes and locations, providing actionable insights to improve public transit efficiency. This research contributes to understanding how transit delays vary across contexts, offering strategies to enhance service reliability for commuters and transit planners."
format: pdf
number-sections: true
toc: true 
bibliography: references.bib
---

@@

# Load Libraries
packages <- c("tidyverse", "kableExtra", "ggplot2", "dplyr", "here", "knitr", "arrow", "randomForest")

load_and_install <- function(package_list) {
  for (x in packages) {
    if (!require(x, character.only = TRUE)) {
      install.packages(i)
      library(x, character.only = TRUE)
    }
  }
}

load_and_install(packages)

# Load Model
model <- readRDS(here::here("models/random_forest_model_delay.rds"))

# Load the model metrics
model_metrics <- readRDS(here::here("models", "evaluation_results.rds"))

# Load Data
cleaned_data <- read_parquet(here::here("data/02-analysis_data/combined_ttc_delay_data.parquet")) %>%
  filter(!is.na(location)) %>%
  mutate(
    location = factor(location, levels = unique(location)),
    day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
    mode = factor(mode, levels = c("Bus", "Streetcar", "Subway"))
  )

```

# Introduction
The performance of public transit systems is a cornerstone of urban mobility, shaping commuter experiences and the broader dynamics of city life. In Toronto, the Toronto Transit Commission (TTC) operates a multimodal network of buses, subways, and streetcars, each of which faces distinct operational challenges. Transit delays remain a persistent issue, disrupting schedules, inconveniencing passengers, and reducing the overall efficiency of the city’s transportation infrastructure. Delays can also erode public trust in transit systems, which is critical for cities aiming to increase public transit use as part of sustainable urban mobility strategies [@citeLitman2023]. Understanding the underlying factors contributing to these delays is essential for designing interventions that enhance system reliability and commuter satisfaction.

This study investigates TTC delays across buses, subways, and streetcars, emphasizing patterns across geographic locations and temporal contexts. Two specific locations—Dundas, in Toronto’s urban core, and Yorkdale, a suburban area—serve as focal points for comparative analysis. Dundas represents a dense, high-demand transit environment often subject to congestion and frequent service interruptions, consistent with findings in other high-density urban transit systems [@citeCats2017]. In contrast, Yorkdale provides a suburban context with less frequent service and lower traffic volumes, which are typically associated with fewer delays and more predictable transit operations [@citeHess2007]. By exploring these contrasting environments, this research aims to uncover delay dynamics shaped by urban density, transit demand, and local infrastructure.

Previous research has examined broad operational inefficiencies, such as traffic congestion and service disruptions, as primary contributors to delays [@citeLitman2023]. However, the complex interactions between transit mode, geographic factors, and temporal patterns remain underexplored [@citeCurrie2012]. This study addresses that gap by leveraging detailed datasets from Open Data Toronto to model delay durations and identify key predictors, including transit mode, travel direction (bound), and day of the week. The findings align with global research highlighting the importance of mode-specific challenges, such as the susceptibility of buses to traffic congestion and the relative reliability of rail-based systems [@citeCats2017; @citeHess2007].

The estimand of this study is delay duration (min_delay), defined as the recorded delay for individual transit trips. Using a Random Forest algorithm, this study models delay duration with predictors such as mode, location, temporal factors, and service regularity (min_gap). The analysis reveals significant variations in delays across transit modes and geographic locations, underscoring the challenges faced by buses in urban settings compared to the more consistent performance of subways. Temporal patterns, such as higher delays during weekdays, further highlight the role of commuter-driven demand in shaping transit reliability.

The structure of this paper is as follows: Section @sec-data details the datasets, cleaning procedures, and key variables used in the analysis. Section @sec-methods outlines the exploratory data analysis and the modeling framework, explaining the rationale for selecting Random Forests. Section @sec-results presents the primary findings, including visualizations and statistical summaries of delay patterns. Finally, Section @sec-discussion interprets the results, explores their implications for TTC operations, and identifies limitations and opportunities for future research.

# Data {#sec-data}
The analysis was conducted using R [@citeR], leveraging the following R packages to facilitate data cleaning, visualization, and analysis: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot2], `dplyr` [@citeDplyr], `here` [@citeHere], `arrow` [@citeArrow], `caret` [@citeCaret], `randomForest` [@citeRandomForest], `stringr` [@citeStringr], `knitr` [@citeKnitr], `testthat` [@citeTestthat] and `readr` [@citeReadr].

Additional guidance for this project was drawn from Telling Stories with Data by Rohan Alexander [@citeTellingStories], which shaped the analytical approach. Details about data cleaning and variable transformations are provided in [@sec-data-cleaning].

## Source of Data
The dataset for this research was obtained from Open Data Toronto, a publicly accessible platform providing detailed municipal datasets. This particular dataset focuses on transit delays across the Toronto Transit Commission (TTC) network, including buses, subways, and streetcars, for the year 2024. The dataset offers a granular view of transit disruptions, capturing key operational metrics and contextual factors. It serves as the primary resource for analyzing delay patterns and identifying the influences of transit mode, geographic location, and temporal factors on service reliability.

The data encompasses several critical variables. ‘mode’ categorizes the transit type—bus, subway, or streetcar—each with distinct operational contexts. The ‘location’ variable specifies where delays occurred, with a focus on Dundas, representing Toronto’s urban core, and Yorkdale, a suburban area. These locations allow for a comparative analysis of transit dynamics across contrasting environments. Temporal variables, such as ‘day_of_week,’ facilitate an understanding of how delays vary between weekdays and weekends, reflecting commuter versus recreational patterns. Directionality, captured in the ‘bound’ variable, enables an exploration of the geographic and directional influences on delays. Finally, the numerical variables ‘min_delay,’ measuring the shortest delay recorded, and ‘min_gap,’ reflecting the time gap between consecutive vehicles, are central to understanding service disruptions and regularity.

This dataset was selected for its level of detail and public accessibility, which align with the research goal of reproducibility and transparency. Alternative datasets, such as internal TTC reports or proprietary transit data, were considered but were unavailable due to licensing restrictions and limited public access. Open Data Toronto, combined with R packages like `opendatatoronto` [@citeOpendatatoronto] and `tidyverse` [@citeTidyverse], enabled seamless integration and analysis.

## Measurement
The TTC delay dataset translates real-world transit disruptions into structured data entries to facilitate an analysis of operational patterns and predictors of service reliability. Key variables include min_delay, min_gap, mode, bound, and location, each capturing distinct dimensions of transit performance. These variables collectively enable a comprehensive examination of delay dynamics across Toronto’s public transit system.

The primary outcome variable, min_delay, represents the delay duration in minutes for each transit trip. This measure reflects the extent of service disruptions, ranging from brief delays to significant interruptions. It serves as a critical metric for assessing the impact of delays on the TTC system's overall reliability. The companion variable, min_gap, quantifies the time gap between consecutive vehicles on the same route, providing insights into service regularity and scheduling efficiency.

The mode variable categorizes the type of transit service—bus, subway, or streetcar—capturing operational differences shaped by infrastructure, traffic interactions, and service management. The bound variable specifies the direction of travel, such as North, South, East, or West, enabling an exploration of geographic and directional influences on delays. Finally, the location variable focuses on key geographic contexts, highlighting patterns in urban (Dundas) and suburban (Yorkdale) settings. This geographic distinction is central to understanding variations in delay dynamics across different transit environments.

Although the dataset does not provide explicit documentation of its collection methods, it is reasonable to infer that the TTC aggregates data through automated monitoring systems, supplemented by manual reporting. Automated data collection likely includes feeds from vehicle tracking systems, GPS technology, and onboard sensors, while manual logs may capture incidents that are not automatically detected. This dual approach ensures a detailed record of service performance, consistent with contemporary transit data practices [@citeLitman2023; @citeCats2017].

The structured nature of the dataset allows for nuanced analyses of delay patterns and their operational drivers. Each variable was selected to capture essential aspects of transit performance, transforming complex real-world phenomena into analyzable components. This structured data foundation enables a systematic exploration of factors influencing TTC delays and offers insights to inform transit planning and operational improvements.

A datasheet is available for viewing through the repository this paper is associated with.

## Outcome Variables
The outcome variable analyzed in this study, min_delay, represents the delay experienced by transit vehicles in minutes. This variable is critical for understanding service disruptions in the Toronto Transit Commission (TTC) system. By modeling min_delay, the analysis aims to quantify delay patterns and identify key factors influencing transit reliability across different modes, geographic locations, and temporal contexts.

### Delay Duration (min_delay)
The min_delay variable measures the recorded delay for a transit trip, expressed in minutes. It serves as a key indicator of service performance, enabling the identification of delay trends and areas requiring operational improvements. As shown in @tbl-min_delay_summary, the summary statistics for min_delay provide insights into the central tendency and variability of transit delays across the dataset.

```{r}
#| label: tbl-min_delay_summary
#| tbl-cap: "Summary statistics for delay duration (min_delay, in minutes)"
#| echo: false
#| warning: false
#| message: false

min_delay_summary <- cleaned_data %>%
  summarise(
    Statistic = c("Shortest Delay (mins)", "Longest Delay (mins)", "Average Delay (mins)", "Median Delay (mins)", "Standard Deviation (mins)"),
    Value = c(
      round(min(min_delay, na.rm = TRUE), 3),
      round(max(min_delay, na.rm = TRUE), 3),
      round(mean(min_delay, na.rm = TRUE), 3),
      round(median(min_delay, na.rm = TRUE), 3),
      round(sd(min_delay, na.rm = TRUE), 3)
    )
  )

knitr::kable(
  min_delay_summary,
  col.names = c("Statistic", "Value"),
  booktabs = TRUE
)

```
The table indicates that min_delay ranges from 0 minutes (indicating no delay) to a maximum of 975 minutes, with a mean delay of 15.614 minutes and a median of 10 minutes. The standard deviation of 44.856 reflects substantial variability in delay durations, suggesting that while many delays are relatively short, there are significant outliers with extreme delay durations.

@fig-min_delay_distribution visualizes the distribution of min_delay using a histogram, highlighting the variable’s central tendency and spread.

```{r}
#| label: fig-min_delay_distribution
#| fig-cap: "Distribution of delay durations (min_delay, in minutes, capped at 100)"
#| echo: false
#| warning: false
#| message: false

# Remove extreme outliers
filtered_data <- cleaned_data %>%
  filter(min_delay <= 100)  # Only include delays of 100 minutes or less

# Plot the histogram
ggplot(filtered_data, aes(x = min_delay)) +
  geom_histogram(binwidth = 5, fill = "#2E86C1", color = "white", alpha = 0.8) +
  labs(
    x = "Delay Duration (minutes)",
    y = "Frequency"
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```
The histogram visualizes the distribution of min_delay after removing extreme outliers (values above 100 minutes). As shown in @fig-min_delay_distribution, the majority of delay durations are concentrated within the range of 0–30 minutes, with a notable peak around the mean and median of approximately 10 minutes. Extreme delays above 100 minutes were excluded for clarity, as these cases are rare and could skew the visualization.

## Predictor Variables
The predictor variables include mode, location, bound, day_of_week, and min_gap. Mode captures the type of transit service—bus, streetcar, or subway—each with distinct operational dynamics influencing delay patterns. Location identifies where delays occur, focusing on DUNDAS, YORKDALE, and OTHER, to explore differences between urban and suburban transit environments. The bound variable specifies the direction of travel—north, south, east, or west—highlighting geographic and directional influences on delays. Temporal factors are represented by day_of_week, allowing an analysis of delay patterns across weekdays and weekends. Finally, min_gap measures the time gap between consecutive vehicles on the same route, reflecting service regularity and its impact on operational performance.

### Mode of Transit
The mode variable captures the type of transit service (Bus, Subway, or Streetcar) and reflects the distinct operational challenges associated with each mode. Buses contend with external factors such as road traffic, weather, and construction, leading to frequent and highly variable delays. Subways are typically influenced by infrastructure issues like signal failures and mechanical disruptions, resulting in relatively shorter and more concentrated delays. Streetcars, operating in mixed traffic, encounter delays from route congestion and shared road usage.

@tbl-mode_delay_summary presents a summary of delay statistics for each transit mode, including average, median, maximum, and minimum delays, along with standard deviations.
```{r}
#| label: tbl-mode_delay_summary
#| tbl-cap: "Summary statistics for delays by transit mode (in minutes)"
#| echo: false
#| warning: false
#| message: false

mode_delay_summary <- cleaned_data %>%
  group_by(mode) %>%
  summarise(
    Average_Delay = round(mean(min_delay, na.rm = TRUE), 3),
    Median_Delay = median(min_delay, na.rm = TRUE),
    Max_Delay = max(min_delay, na.rm = TRUE),
    Min_Delay = min(min_delay, na.rm = TRUE),
    Std_Dev = round(sd(min_delay, na.rm = TRUE), 3)
  )

knitr::kable(
  mode_delay_summary,
  col.names = c("Transit Mode", "Average Delay", "Median Delay", "Max Delay", "Min Delay", "Std Dev"),
  booktabs = TRUE
)

```
As shown in @tbl-mode_delay_summary, buses experience the highest average delay (21.273 minutes) and the greatest variability, with a standard deviation of 54.51 minutes. Streetcars follow with an average delay of 15.477 minutes and a slightly lower standard deviation of 32.713 minutes. Subways show the lowest average delay of just 2.806 minutes, underscoring the relative reliability of rail-based transit compared to road-based services. These findings highlight the disproportionate impact of external factors like road congestion on buses and streetcars.

### Geographic Location
The location variable provides a geographic lens for analyzing delays, focusing on two key areas: Dundas (an urban, high-traffic area) and Yorkdale (a suburban area with less frequent service). This variable allows for a comparison of transit performance in urban versus suburban contexts, offering insights into how population density and service frequency affect delays.

@tbl-location_summary summarizes delay statistics for each location, capturing key metrics like average delay, median delay, and variability.
\newpage
```{r}
#| label: tbl-location_summary
#| tbl-cap: "Summary statistics for the predictor variable (location)"
#| echo: false
#| warning: false
#| message: false

location_delay_summary <- cleaned_data %>%
  group_by(location) %>%
  summarise(
    Average_Delay = round(mean(min_delay, na.rm = TRUE), 3),
    Median_Delay = median(min_delay, na.rm = TRUE),
    Max_Delay = max(min_delay, na.rm = TRUE),
    Min_Delay = min(min_delay, na.rm = TRUE),
    Std_Dev = round(sd(min_delay, na.rm = TRUE), 3)
  )

knitr::kable(
  location_delay_summary,
  col.names = c("Location", "Average Delay", "Median Delay", "Max Delay", "Min Delay", "Std Dev"),
  booktabs = TRUE
)

```
According to @tbl-location_summary, Dundas exhibits slightly higher average delays (13.491 minutes) than Yorkdale (11.521 minutes), with a standard deviation of 31.966 minutes compared to Yorkdale's 23.821 minutes. This suggests that urban environments like Dundas experience more variable and longer delays due to higher transit demand and traffic density. In contrast, Yorkdale benefits from lower demand and more predictable traffic conditions.

### Directionality
The bound variable identifies the travel direction (North, South, East, or West) of delayed transit vehicles. This dimension helps uncover localized bottlenecks or infrastructure-related issues that may disproportionately affect certain routes.

@tbl-bound_delay_summary provides a detailed breakdown of delay statistics by direction.
```{r}
#| label: tbl-bound_delay_summary
#| tbl-cap: "Summary statistics for delays by directional bound (in minutes)"
#| echo: false
#| warning: false
#| message: false

bound_delay_summary <- cleaned_data %>%
  filter(!is.na(bound)) %>% # Exclude rows where 'bound' is NA
  group_by(bound) %>%
  summarise(
    Average_Delay = round(mean(min_delay, na.rm = TRUE), 3),
    Median_Delay = median(min_delay, na.rm = TRUE),
    Max_Delay = max(min_delay, na.rm = TRUE),
    Min_Delay = min(min_delay, na.rm = TRUE),
    Std_Dev = round(sd(min_delay, na.rm = TRUE), 3)
  )

knitr::kable(
  bound_delay_summary,
  col.names = c("Direction", "Average Delay", "Median Delay", "Max Delay", "Min Delay", "Std Dev"),
  booktabs = TRUE
)

```
As shown in @tbl-bound_delay_summary, delays are relatively consistent across directions, with average delays ranging from 14.742 minutes (South) to 17.252 minutes (North). However, the standard deviation varies across directions, with the Northbound direction showing the greatest variability (49.128 minutes). These findings may reflect differences in route infrastructure, traffic bottlenecks, or operational scheduling.

### Temporal Factors
Temporal factors, captured by the day of the week, provide insights into how commuter and recreational travel patterns influence transit delays. Weekday delays typically reflect peak commuter traffic, while weekend delays may indicate reduced service levels or unique traffic patterns tied to recreational activities.

@tbl-dow_delay_summary presents the summary statistics for delays by day of the week, highlighting differences in average delay, variability, and extreme values.
```{r}
#| label: tbl-dow_delay_summary
#| tbl-cap: "Summary statistics for delays by day of the week (in minutes)"
#| echo: false
#| warning: false
#| message: false

dow_delay_summary <- cleaned_data %>%
  group_by(day_of_week) %>%
  summarise(
    Average_Delay = round(mean(min_delay, na.rm = TRUE), 3),
    Median_Delay = median(min_delay, na.rm = TRUE),
    Max_Delay = max(min_delay, na.rm = TRUE),
    Min_Delay = min(min_delay, na.rm = TRUE),
    Std_Dev = round(sd(min_delay, na.rm = TRUE), 3)
  )

knitr::kable(
  dow_delay_summary,
  col.names = c("Day of Week", "Average Delay", "Median Delay", "Max Delay", "Min Delay", "Std Dev"),
  booktabs = TRUE
)

```
According to @tbl-dow_delay_summary, average delays are relatively consistent across weekdays, peaking slightly on Monday (15.916 minutes) and Friday (15.649 minutes). The standard deviations for weekdays, such as Monday (48.686 minutes) and Friday (45.056 minutes), indicate substantial variability in delays during peak commuter periods. Weekend delays are slightly lower on average, with Saturday showing the lowest average delay (14.748 minutes) and reduced variability (39.019 minutes). These patterns align with the expected reduction in transit demand and traffic congestion during weekends.

### Time Gap Between Vehicles (min_gap)
The min_gap variable captures the time gap (in minutes) between consecutive vehicles on the same route, providing insight into service regularity. Maintaining consistent gaps is crucial for operational efficiency and minimizing passenger wait times. @tbl-min_gap_summary presents the summary statistics for min_gap.
```{r}
#| label: tbl-min_gap_summary
#| tbl-cap: "Summary statistics for time gaps between vehicles (min_gap, in minutes)"
#| echo: false
#| warning: false
#| message: false

min_gap_summary <- cleaned_data %>%
  summarise(
    Statistic = c("Shortest Gap (mins)", "Longest Gap (mins)", "Average Gap (mins)", "Median Gap (mins)", "Standard Deviation (mins)"),
    Value = c(
      round(min(min_gap, na.rm = TRUE), 3),
      round(max(min_gap, na.rm = TRUE), 3),
      round(mean(min_gap, na.rm = TRUE), 3),
      round(median(min_gap, na.rm = TRUE), 3),
      round(sd(min_gap, na.rm = TRUE), 3)
    )
  )

knitr::kable(
  min_gap_summary,
  col.names = c("Statistic", "Value"),
  booktabs = TRUE
)

```
The data indicates that min_gap ranges from 0 to 997 minutes, with an average time gap of 24.438 minutes and a median of 19 minutes. A standard deviation of 47.659 reflects substantial variability, likely influenced by external factors such as traffic congestion, irregular vehicle arrivals, and operational inefficiencies. The presence of extremely high values, such as 997 minutes, suggests rare but significant disruptions that may warrant further investigation or exclusion in detailed analyses.

@fig-min_gap_distribution depicts the distribution of min_gap, using a histogram to illustrate common intervals and variability.
```{r}
#| label: fig-min_gap_distribution
#| fig-cap: "Distribution of time gaps between vehicles (min_gap, in minutes)"
#| echo: false
#| warning: false
#| message: false

ggplot(cleaned_data, aes(x = min_gap)) +
  geom_histogram(binwidth = 10, fill = "#FF5733", color = "white", alpha = 0.8) +
  labs(
    x = "Time Gap Between Vehicles (minutes)",
    y = "Frequency"
  ) +
  scale_x_continuous(limits = c(0, 200), breaks = seq(0, 200, by = 20)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```
The histogram illustrates that the majority of min_gap values fall within the 15–30 minute range, with a noticeable peak around 20 minutes. While most values are clustered within this interval, there are occasional larger gaps that extend up to 100 minutes, reflecting variability in service intervals. These deviations may highlight operational inconsistencies or external disruptions affecting transit schedules.

# Methods {#sec-methods}
This section outlines the methodology employed to analyze patterns in Toronto Transit Commission (TTC) delays (min_delay) using a Random Forest model. By leveraging an ensemble approach, the Random Forest effectively handles nonlinear relationships and interactions between predictors, providing robust predictions for delay durations across diverse transit contexts.

## Model Overview
The goal of the model is to predict the delay duration (min_delay) in minutes for TTC transit vehicles using the following predictors:

$$
\text{min\_delay} \sim \text{mode} + \text{location} + \text{bound} + \text{day\_of\_week} + \text{min\_gap}\tag{1}
$$

Where:

- **mode**: The transit mode (Bus, Streetcar, or Subway).
- **location**: Geographic location (Dundas, Yorkdale, or Other).
- **bound**: Travel direction (North, South, East, or West).
- **day_of_week**: Day of the week (Monday through Sunday).
- **min_gap**: Time gap between vehicles in minutes, representing service regularity.

## Training and Validation
To ensure robust performance, the dataset was split into training (80%) and testing (20%) sets using stratified sampling to maintain the distribution of 
min_delay
min_delay. The training set was used to build the Random Forest model, while the testing set evaluated its predictive accuracy.

The model was implemented in R using the randomForest package, with the number of trees (`ntree`) set to 500. This parameter balances computational efficiency and model stability by ensuring sufficient averaging across trees.
$$
\hat{y}_i = \frac{1}{T} \sum_{t=1}^{T} f_t(\mathbf{x}_i)\tag{2}
$$

Where:

- $\hat{y}_i$: Predicted delay for observation $i$.
- $T$: Total number of trees ($T = 500$).
- $f_t(\mathbf{x}_i)$: Prediction from tree $t$ based on predictors $\mathbf{x}_i$.

The model hyperparameters, including the default number of predictors sampled at each split ($\sqrt{p}$, where $p$ is the number of predictors), were kept unchanged, as they offered a balanced trade-off between bias and variance.

## Evaluation Metrics
The model’s performance was assessed using:

- **Root Mean Squared Error (RMSE):**
$$
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^{n} \left(\hat{y}_i - y_i\right)^2}\tag{3}
$$
Where $n$ is the number of observations, $y_i$ is the observed delay, and $\hat{y}_i$ is the predicted delay.

- **Coefficient of Determination ($R^2$):**
$$
R^2 = 1 - \frac{\sum_{i=1}^{n} \left(y_i - \hat{y}_i\right)^2}{\sum_{i=1}^{n} \left(y_i - \bar{y}\right)^2}\tag{4}
$$
This measures the proportion of variance in the observed data explained by the model. Values close to 1 indicate strong predictive performance.

## Model Justification, Assumptions, and Limitations
The Random Forest regression model was selected for its ability to handle the diverse and nonlinear relationships within the dataset, making it the most suitable choice for analyzing TTC delays. This model is particularly effective at capturing complex interactions between predictors, such as transit mode and day of the week, which influence delay durations.

The justification for using Random Forest lies in its ability to model nonlinear relationships without assuming specific functional forms. Unlike linear regression, which oversimplifies these relationships, Random Forest accommodates interactions between predictors and provides robust predictions even in the presence of multicollinearity. For example, the influence of min_gap on delays is not consistent across different transit modes, a pattern that Random Forest can identify and model effectively. Additionally, the algorithm provides feature importance metrics, offering interpretable insights into the drivers of delays, such as mode, location, and day_of_week.

Alternative models, such as linear regression, were considered but found to be insufficient for this analysis. Linear regression assumes linear relationships and independence between predictors, leading to poor performance metrics, including lower $R^2$ values and higher RMSE scores, when applied to this dataset. These results highlighted the need for a more flexible model, prompting the selection of Random Forest.

Key predictors were evaluated for their importance using the Mean Decrease in Accuracy (MDA) metric, which measures the reduction in predictive accuracy when a variable is excluded. This analysis, performed using the `randomForest` package [@citeRandomForest], revealed that min_gap (time between consecutive vehicles) and mode (transit type) were the most significant predictors. The prominence of min_gap underscores its role in service regularity, while the importance of mode reflects the operational differences among buses, subways, and streetcars. Other influential predictors included day_of_week and location, capturing temporal and geographic dimensions of delays.

While Random Forest excelled in its predictive capabilities, the model operates under several assumptions and faces certain limitations. First, although it handles outliers better than simpler models, extreme values, such as exceptionally high min_gap durations, can still affect individual tree predictions. These outliers were mitigated during preprocessing (see [@sec-data-cleaning]). Second, the model is less interpretable compared to linear regression, as it does not provide straightforward coefficients for predictors. Feature importance metrics were used to partially address this limitation. Third, Random Forest can be sensitive to imbalances in predictor distributions, such as fewer observations for specific locations or directional bounds. Careful data cleaning and preprocessing steps were implemented to reduce this bias.

Despite its strengths, the model has limitations in its applicability. For example, it may struggle with extreme imbalances in predictor distributions or scenarios requiring highly interpretable predictions. Future work could explore ensemble methods that integrate Random Forest with more interpretable models to address these challenges. The comprehensive evaluation and selection process affirm the Random Forest model as the best fit for understanding and predicting TTC delays.

See [@sec-model-details] for additional details.

# Results {#sec-results}
## Visualizing Delays

@fig-delays_by_location illustrates the average delay duration across three key Toronto locations: Dundas, Yorkdale, and "Other." The figure shows distinct variations in delay durations, with Other locations experiencing the longest delays, averaging 15.7 minutes, compared to 13.5 minutes at Dundas and 11.5 minutes at Yorkdale. These results suggest that factors specific to "Other" locations, such as less efficient transit infrastructure or increased congestion, contribute to higher delays. Conversely, Yorkdale exhibits relatively shorter delays, which could reflect better-managed transit operations or reduced traffic volume in the area.
```{r}
#| label: fig-delays_by_location
#| fig-cap: "Average delays across Toronto locations, with average delay values annotated."
#| echo: false
#| warning: false
#| message: false

avg_delay_by_location <- cleaned_data %>%
  group_by(location) %>%
  summarise(avg_delay = mean(min_delay, na.rm = TRUE))

ggplot(avg_delay_by_location, aes(x = location, y = avg_delay, fill = location)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_text(aes(label = round(avg_delay, 1)), vjust = 2, size = 4) +
  labs(
    x = "Location",
    y = "Average Delay Duration (minutes)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

```

@fig-temporal_trends_adjusted highlights the temporal trends in average TTC delays across the days of the week. Mondays show the highest average delays (15.9 minutes), potentially due to increased traffic demand as the workweek begins. Saturdays see the lowest average delays (14.7 minutes), aligning with reduced commuter traffic during weekends. While the variations in delays across days are relatively small, these findings emphasize the subtle but consistent influence of weekly transit demand patterns on delay durations.
```{r}
#| label: fig-temporal_trends_adjusted
#| fig-cap: "Temporal trends in TTC delays, with average delay values annotated."
#| echo: false
#| warning: false
#| message: false

avg_delay_by_day <- cleaned_data %>%
  group_by(day_of_week) %>%
  summarise(avg_delay = mean(min_delay, na.rm = TRUE))

ggplot(avg_delay_by_day, aes(x = day_of_week, y = avg_delay, group = 1)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_point(color = "red", size = 3) +
  geom_text(aes(label = round(avg_delay, 1)), vjust = 1.3, hjust = 1.4, size = 3) +
  labs(
    x = "Day of the Week",
    y = "Average Delay Duration (minutes)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

The distribution of delays across different transit modes—bus, streetcar, and subway—is presented in @fig-delay_by_mode. Buses exhibit the most variability, with delays ranging from a few minutes to over 750 minutes, likely due to their exposure to road traffic and weather conditions. Streetcars show moderate delays, reflecting their shared use of road infrastructure but also benefiting from dedicated tracks in certain areas. Subways demonstrate the most consistent performance, with shorter and less variable delays, likely due to their isolated underground operation.
```{r}
#| label: fig-delay_by_mode
#| fig-cap: "Distribution of delays by mode of transit"
#| echo: false
#| warning: false
#| message: false

ggplot(cleaned_data, aes(x = mode, y = min_delay, color = mode)) +
  geom_jitter(width = 0.3, alpha = 0.6, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4, shape = 23, fill = "gold") +
  labs(
    x = "Mode of Transit",
    y = "Delay Duration (minutes)",
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

```

@fig-delays_by_direction examines the average delays based on the travel direction: East, West, North, and South. Northbound trips have the longest average delays (17.3 minutes), while southbound trips experience the shortest delays (14.7 minutes). East and westbound trips have comparable delays, averaging 16.6 minutes and 16.4 minutes, respectively. These findings may reflect route-specific operational challenges or variations in traffic and transit infrastructure across different directions.
```{r}
#| label: fig-delays_by_direction
#| fig-cap: "Average delays by travel direction."
#| echo: false
#| warning: false
#| message: false

avg_delay_by_direction <- cleaned_data %>%
  filter(!is.na(bound)) %>% # Exclude rows where 'bound' is NA
  group_by(bound) %>%
  summarise(avg_delay = mean(min_delay, na.rm = TRUE))

ggplot(avg_delay_by_direction, aes(x = bound, y = avg_delay, fill = bound)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_text(aes(label = round(avg_delay, 1)), vjust = 2, size = 4) +
  labs(
    x = "Travel Direction",
    y = "Average Delay Duration (minutes)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

```

The scatterplot in @fig-gap_vs_delay illustrates the relationship between the time gap between consecutive transit vehicles and the resulting delay duration. The trend line indicates a positive correlation: larger time gaps correspond to longer delays. This relationship highlights the importance of maintaining consistent headways to minimize delays and optimize transit performance. The scatterplot also shows a wide range of delays for small time gaps, suggesting that other factors, such as operational issues or traffic conditions, play a significant role in influencing delays.
```{r}
#| label: fig-gap_vs_delay
#| fig-cap: "Relationship between time gap and delay duration."
#| echo: false
#| warning: false
#| message: false

ggplot(cleaned_data, aes(x = min_gap, y = min_delay)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +
  labs(
    x = "Time Gap (minutes)",
    y = "Delay Duration (minutes)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )

```

A closer examination of streetcar lines 501 and 510, shown in @fig-streetcar_lines, reveals notable differences in their delay durations. Line 501 experiences longer average delays (16.1 minutes) compared to line 510 (10.7 minutes). These disparities may be attributed to differences in route length, traffic congestion, or infrastructure quality. Line 501’s longer delays suggest the need for targeted interventions to improve reliability along this route.
```{r}
#| label: fig-streetcar_lines
#| fig-cap: "Average delays for Streetcar Lines 501 and 510."
#| echo: false
#| warning: false
#| message: false

streetcar_delays <- cleaned_data %>%
  filter(mode == "Streetcar" & line %in% c("501", "510")) %>%
  group_by(line) %>%
  summarise(avg_delay = mean(min_delay, na.rm = TRUE))

ggplot(streetcar_delays, aes(x = line, y = avg_delay, fill = line)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_text(aes(label = round(avg_delay, 1)), vjust = 2, size = 4) +
  labs(
    x = "Streetcar Line",
    y = "Average Delay Duration (minutes)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

```

@fig-day_date_trends explores the interplay between daily delays and the day of the week over time. The temporal trends reveal significant variations, with some months exhibiting consistent spikes in delays on specific weekdays. For instance, delays are particularly high during April, potentially reflecting seasonal disruptions such as weather-related challenges or scheduled maintenance. These patterns emphasize the need for adaptive scheduling and operational strategies to address temporal fluctuations in delay durations.
```{r}
#| label: fig-day_date_trends
#| fig-cap: "Temporal Trends in Delays by Day of the Week and Date."
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 6

temporal_delays <- cleaned_data %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d")) %>%
  group_by(date, day_of_week) %>%
  summarise(avg_delay = mean(min_delay, na.rm = TRUE)) %>%
  ungroup()
